---
title: "Make metadata"
author: "Jennie Yao"
date: "2023-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(here)
library(data.table)
```

**NOTE: THIS SCRIPT NO LONGER WORKS, FILE FORMATS HAVE BEEN CHANGED. TO ADD NEW SAMPLES, MANUALLY ADD THEM TO `/Neoantigen ML project/data/meta_manual.csv`**



Output directory

```{r set output directory}
outdir <- here("output", "01_make_meta.Rmd")
fs::dir_create(outdir)
```




# Meta data requirements

Mata data for each patient (make each row as one patient) 
Columns to include:  

* patient id -- use id that match the Immuno Pipeline and design tracking sheet (need some clean up)
* Cancer type of the patient (get from Immuno Pipeline and design tracking sheet)
* Known gene of that cancer (from CancerGeneCensus-Mar2023.tsv) -- manual enter
* Purity estimate (?? need directions on how to get this)  
  - biopsy from tumor is mixed of normal and cancer cells
  - how much of the biopsy sample is tumor cells
  - purity = VAF of driver gene * 2
  - need a list of driver gene for cancer type
  - if patient has KRAS mutation, has pancrease cancer, we take the patient's KRAS VAF *2
* patient's ITB file path  
* patient's all_epitopes path  
* patient's class 2 path  
* whether this patient was included in the training data/testing data (?? patient peptides will be combined and then make training and testing set)  -- divide based on patients 
* total number of peptides of the given patient (number of rows in the ITB review file)  
* number of "Pass" peptides in the given patient (count from "Tier" column in the ITB review file)  
* number of "Accept" peptides in the given patient (get from "Evaluation" column in the ITB review file)  
* basic stats


Other files
itb_review:
smb://storage1.ris.wustl.edu/gillandersw/Active/Project_0001_Clinical_Trials/CTEP/analysis/TWJF-10146-MO011-0021/itb-review-files/10146-0021.Annotated.Neoantigen_Candidates.tsv
all_epitopes:
smb://storage1.ris.wustl.edu/gillandersw/Active/Project_0001_Clinical_Trials/CTEP/analysis/TWJF-10146-MO011-0021/gcp_immuno/final_results/pVACseq/mhc_i/TWJF-10146-0021-Tumor_Lysate.all_epitopes.tsv
class2:
/Volumes/gillandersw/Active/Project_0001_Clinical_Trials/CTEP/analysis/TWJF-10146-MO011-0021/gcp_immuno/final_results/pVACseq/mhc_ii/TWJF-10146-0021-Tumor_Lysate.all_epitopes.aggregated.tsv

pvacseq-v4 updated pipeline - latest rc
all_epitopes: 
/Volumes/gillandersw/Active/Project_0001_Clinical_Trials/pancreas_leidos/analysis/TWJF-5120-07_BGA-2109/pvacseq-v4/result-median-score-rc11/MHC_Class_I/TWJF-5120-07-5120-07_tumor_lysate.all_epitopes.tsv
class2:
/Volumes/gillandersw/Active/Project_0001_Clinical_Trials/pancreas_leidos/analysis/TWJF-5120-07_BGA-2109/pvacseq-v4/result-median-score-rc11/MHC_Class_II/TWJF-5120-07-5120-07_tumor_lysate.all_epitopes.aggregated.tsv


## Modify from an existing patient file 

Read reference metadata file (may change, this file has all the patient ID for this study)
```{r read ref meta, message=FALSE}
ref_meta <- read_xlsx(here("data", "Immuno Pipeline and Design Tracking.xlsx"))

# Remove the first rows of comments
ref_meta1 <- ref_meta[-1, ]

# Make the row now as column names
colnames(ref_meta1) <- ref_meta1[1, ]

# Remove this row (which is now the column names)
meta1 <- ref_meta1[-1, ]
```

Clean the dataframe, make a extra column that is the `patient ID`
```{r meta extra columns}
meta2 <- meta1 %>% 
  mutate(`N (drive link)` = as.integer(`N (drive link)`)) %>%
  mutate(patient_id = ifelse(
      startsWith(`Case (Box link)`, "JLF"),
      gsub(".*\\((.*?)\\).*", "\\1", `Case (Box link)`),
      `Case (Box link)`
    )) %>%
  mutate(patient_id = ifelse(startsWith(patient_id, "5120"), 
                           gsub("_", "-", patient_id), 
                           patient_id))
```


## Add peptide count, pass count, and accept count columns to the metadata

Write a function to add a column in the metadata file that counts the number of peptides for each patient, number of passing peptides, numbe of accepted peptides

the above information will be pulled from the itb_review folder containing all ITB review files for each patient
```{r get patient itb files}
# list all available patient ITB files in the folder
patient_itb_path <- list.files(here("data", "itb_review"),  full.names = TRUE) 
# extract a list of patient id from file name (NOTE: not all patients in metadata file are in the folder)
patient_id_itb <- patient_itb_path %>% 
  basename() %>% 
  gsub(".revd.Annotated.Neoantigen_Candidates.tsv$", "", .)
```

Define function

* input: a patient's itb file path; metadata that we want to modify
* output: a dataframe with the number of peptides, the number of pass, number of accept for a given patient's file

```{r function for adding peptide counts, eval=FALSE, include=FALSE}
extract_num_peptides <- function(patient_itb_path){
  # extract patient id from itb file path
  patient_id_itb <- patient_itb_path %>% 
    basename() %>% 
    gsub(".revd.Annotated.Neoantigen_Candidates.tsv$", "", .)
  
  # read the patient's itb file
  itb_data <- read.table(patient_itb_path, header = TRUE, sep = "\t")
  
  n_peptides <- nrow(itb_data)
  n_pass <- sum(itb_data$Tier == "Pass")
  n_accept <- sum(itb_data$Evaluation == "Accept")
  
  # patient ID in metadata are in the form 5120_06, but in file names it is 5120-06
  #patient_id_itb <- ifelse(startsWith(patient_id_itb, "5120"), 
                           #gsub("-", "_", patient_id_itb), 
                           #patient_id_itb)
  
  data.frame(patient_id = patient_id_itb, n_peptides, n_pass, n_accept)
  
}
```

```{r add peptide counts columns to meta, eval=FALSE, include=FALSE}
# Use lapply to process each file and store the results in a list
itb_patient_list <- lapply(patient_itb_path, extract_num_peptides)

# Merge the list of data frames into a single dataframe
itb_peptide_counts <- do.call(rbind, itb_patient_list)

# join columns of counts to the metadata
meta3 <- left_join(meta2, itb_peptide_counts, by = "patient_id")
```


## Add file paths

```{r get all file paths}
# list all available patient ITB files in the folder
patient_itb_path <- list.files(here("data", "itb_review"),  full.names = TRUE) 
# list patients that have available itb_review files
patient_id_itb <- patient_itb_path %>% 
  basename() %>% 
  gsub(".revd.Annotated.Neoantigen_Candidates.tsv$", "", .) %>% 
  gsub(".Annotated.Neoantigen_Candidates.tsv$", "", .)

# list all available patient all_epitopes files in the folder
patient_epi_path <- list.files(here("data", "all_epitopes"),  full.names = TRUE) 

# list paths to all class 2 files
patient_c2_path <- list.files(here("data", "class2"),  full.names = TRUE) 
```

Define a function that out all file paths to a dataframe
input: patient_id, project directory where there is a sub-folder called "data" that contains three sub-folders named "itb_review", "all_epitopes", "class2"
output: a dataframe with the file paths of the patient's file paths (3 columns for 3 file types)
```{r function for get file paths}
get_file_paths <- function(patient_id, project_dir){
  # NOTE: the patient_id as input is the format in the file names, not in metadata
  
  # construct the directory paths for each file type
  itb_path <- file.path(project_dir, "data", "itb_review")
  epi_path <- file.path(project_dir, "data", "all_epitopes")
  c2_path <- file.path(project_dir, "data", "class2")
  
  # get the file from the patient_id
  itb_file_path <- list.files(itb_path, full.names = TRUE) %>% 
    grep(patient_id, ., value = TRUE)
  epi_file_path <- list.files(epi_path, full.names = TRUE) %>% 
    grep(patient_id, ., value = TRUE)
  c2_file_path <- list.files(c2_path, full.names = TRUE) %>% 
    grep(patient_id, ., value = TRUE)
  
  # return a dataframe with patient_id, and file paths
  data.frame(patient_id, 
             itb_review_path = itb_file_path, 
             all_epitopes_path = epi_file_path, 
             class2_path = c2_file_path)
}

# test 
get_file_paths("5120-06", here())
```

```{r add file paths columns to meta}
# Use lapply to process each file and store the results in a list
file_paths_df <- lapply(patient_id_itb, get_file_paths, here()) %>% 
  do.call(rbind, .)

# join columns of counts to the metadata
meta4 <- left_join(meta2, file_paths_df, by = "patient_id")
meta4
```

## Save final data

```{r eval=FALSE, include=FALSE}
write.csv(meta4, file=here(outdir, "metadata.csv"), row.names = FALSE)
```



`metadata.csv` is reviewed manually to modify patient_id, driver genes
