---
title: "Cleaning for prediction"
author: "Jennie Yao"
date: "2023-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(tidyverse)
library(readxl)
library(here)
library(data.table)
library(randomForest)
library(openxlsx) # save excel file
library(parallel)
library(missForest) # random forest missing value imputation
```

Output directory
```{r set output directory}
outdir <- here("output", "05-2_cleaning_prediction.Rmd")
fs::dir_create(outdir)
```

**NOTE: This file is for external validation/prediciton samples only! Not used during training (for training use script 05_cleaning_imputation.Rmd)**

Run this after `04_reduce_columns_for_training.Rmd`

Goal: 
* uniform the column names as done in `05_cleaning_imputation.Rmd` 
* save each patient data separately for prediction


OUTCOME: 
Output file: `patient_id_cleaned.RDS`



Read metadata
```{r read metadata}
meta <- fread(here("output", "02_make_meta2.Rmd", "metadata_count_purity.csv"))
```

get a list of patient_id that we will make predictions 
```{r extract patient list}
patient_id_list <- meta %>% 
  filter(external_validation == "Yes") %>% 
  select(patient_id) %>% 
  pull(patient_id)
```


```{r}
col_names_keep1 <- readRDS(file = file.path(here("output", "05_cleaning_imputation.Rmd"), "col_names_keep1.RDS"))
col_names_keep2 <- readRDS(file = file.path(here("output", "05_cleaning_imputation.Rmd"), "col_names_keep2.RDS"))
levels_list <- readRDS(file = file.path(here("output", "05_cleaning_imputation.Rmd"), "levels_list.RDS"))

# Function to process each file and extract ID
unify_columns <- function(patient_id_test, in_dir, out_dir) {
  # read files through patient id
  reduced_file_path <-
    list.files(in_dir, 
               pattern = paste0("^.*", patient_id_test, ".*\\.RDS$"), 
               full.names = TRUE) 
  reduced_data <- readRDS(reduced_file_path) %>% 
    mutate(patient_id = patient_id_test)
  
  # modify column names
  if (!any(colnames(reduced_data) %in% col_names_keep1)) {
    reduced_data <- reduced_data %>%
      rename(
        "NetMHCpanEL.WT.IC50.Score" = "NetMHCpanEL.WT.Score",
        "NetMHCpanEL.MT.IC50.Score" = "NetMHCpanEL.MT.Score",
      )
  }
  
  if (!any(colnames(reduced_data) %in% col_names_keep2)) {
    reduced_data <- reduced_data %>%
      rename(
        "MHCflurry.WT.IC50.Score" = "MHCflurry.WT.Score",
        "MHCflurry.MT.IC50.Score" = "MHCflurry.MT.Score",
        "MHCnuggetsI.WT.IC50.Score" = "MHCnuggetsI.WT.Score",
        "MHCnuggetsI.MT.IC50.Score" = "MHCnuggetsI.MT.Score",
        "NetMHC.WT.IC50.Score" = "NetMHC.WT.Score", 
        "NetMHC.MT.IC50.Score" = "NetMHC.MT.Score",
        "NetMHCcons.WT.IC50.Score" = "NetMHCcons.WT.Score", 
        "NetMHCcons.MT.IC50.Score" = "NetMHCcons.MT.Score", 
        "NetMHCpan.WT.IC50.Score" = "NetMHCpan.WT.Score", 
        "NetMHCpan.MT.IC50.Score" = "NetMHCpan.MT.Score", 
         
        "PickPocket.WT.IC50.Score" = "PickPocket.WT.Score", 
        "PickPocket.MT.IC50.Score" = "PickPocket.MT.Score", 
        "SMM.WT.IC50.Score" = "SMM.WT.Score", 
        "SMM.MT.IC50.Score" = "SMM.MT.Score",
        "SMMPMBEC.WT.IC50.Score" = "SMMPMBEC.WT.Score", 
        "SMMPMBEC.MT.IC50.Score" = "SMMPMBEC.MT.Score"
      )
  }
  
  # change chr columns to factor levels
  reduced_data2 <- reduced_data %>% 
    #select(-Percentile.Fail) %>% 
    mutate(
        Prob.match = factor(Prob.match, levels = 0:1, labels = c("NO", "YES")),
        Ref.Match = factor(Ref.Match, levels = 0:1, labels = c("FALSE", "TRUE")),
        TSL = as.factor(TSL),
        TSL = factor(TSL, levels = levels_list$TSL),
        driver_gene = factor(driver_gene, levels = 0:1, labels = c("NO", "YES")),
        #Evaluation = factor(Evaluation, levels = unique(Evaluation)), 
        Biotype = factor(Biotype),
        Biotype = factor(Biotype, levels = levels_list$Biotype),
        Variant.Type = factor(Variant.Type),
        Variant.Type = factor(Variant.Type, levels = levels_list$Variant.Type)
      )
  
  # impute some missing values 
  reduced_data_final <- reduced_data2 %>%
  mutate(
    NetMHC.WT.IC50.Score = if_else(is.na(NetMHC.WT.IC50.Score), IC50.WT.class1, NetMHC.WT.IC50.Score),
    SMM.WT.IC50.Score = if_else(is.na(SMM.WT.IC50.Score), IC50.WT.class1, SMM.WT.IC50.Score),
    SMMPMBEC.WT.IC50.Score = if_else(is.na(SMMPMBEC.WT.IC50.Score), IC50.WT.class1, SMMPMBEC.WT.IC50.Score),
    NetMHC.MT.IC50.Score = if_else(is.na(NetMHC.MT.IC50.Score), IC50.MT.class1, NetMHC.MT.IC50.Score),
    SMM.MT.IC50.Score = if_else(is.na(SMM.MT.IC50.Score), IC50.MT.class1, SMM.MT.IC50.Score),
    SMMPMBEC.MT.IC50.Score = if_else(is.na(SMMPMBEC.MT.IC50.Score), IC50.MT.class1, SMMPMBEC.MT.IC50.Score),
    NetMHC.WT.Percentile = if_else(is.na(NetMHC.WT.Percentile), percentile.WT.class1, NetMHC.WT.Percentile),
    SMM.WT.Percentile = if_else(is.na(SMM.WT.Percentile), percentile.WT.class1, SMM.WT.Percentile),
    SMMPMBEC.WT.Percentile = if_else(is.na(SMMPMBEC.WT.Percentile), percentile.WT.class1, SMMPMBEC.WT.Percentile),
    NetMHC.MT.Percentile = if_else(is.na(NetMHC.MT.Percentile), percentile.MT.class1, NetMHC.MT.Percentile),
    SMM.MT.Percentile = if_else(is.na(SMM.MT.Percentile), percentile.MT.class1, SMM.MT.Percentile),
    SMMPMBEC.MT.Percentile = if_else(is.na(SMMPMBEC.MT.Percentile), percentile.MT.class1, SMMPMBEC.MT.Percentile)
  )
  
  # Apply missForest to impute missing values in training set
  # Exclude, ID, patient_id, and Evaluation columns during imputation
  set.seed(918)
  imputed_df <- missForest(reduced_data_final[, 4:ncol(reduced_data_final)-1])
  
  # Extract the imputed dataset
  imputed_df2 <- imputed_df$ximp
  
  # paste the ID, patient_id back to the imputed value
  columns_to_join <- c("ID", "patient_id", "Evaluation")
  df_final <- bind_cols(reduced_data_final[columns_to_join], imputed_df2)
  
  
  # save cleaned and imputed data
  out_name_rds <- paste0(patient_id_test, "_", "cleaned.RDS")
  saveRDS(df_final, file = file.path(out_dir, out_name_rds))
  
  # write a excel version
  out_name_xl <- paste0(patient_id_test, "_", "cleaned.xlsx")
  write.xlsx(df_final, file.path(outdir, out_name_xl), rowNames = FALSE)

  
  df_final
}
```

```{r}
# Apply the function to each file and combine the results into a list of dataframes
#df_list <- lapply(
  #patient_id_list,
  #unify_columns,
  #in_dir = file.path(here(), "output", "04_reduce_columns_for_training.Rmd"),
  #out_dir = outdir
#)

df_list <- mclapply(
  patient_id_list,
  unify_columns,
  in_dir = file.path(here(), "output", "04_reduce_columns_for_training.Rmd"),
  out_dir = outdir,
  mc.cores = 4
)
```



