---
title: "Wrong predictions analysis after threshold adjustment"
author: "Jennie Yao"
date: "2024-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(data.table)
library(tidyverse)
library(here)
library(openxlsx)
library(gridExtra)
library(ggpubr)
```


Output directory
```{r set output directory}
outdir <- here("output", "08-3_prediction_analysis.Rmd")
fs::dir_create(outdir)
```

After adjusting the threshold for predicitons (`08_randomForest_predict.Rmd`) and looking at the performance confusion matrix (`08-1_threshold_adj.Rmd`), which was an iterative process, The new threshold produced some predicitons in which some peptides were predicted as `Reject` but in fact should have been `Review`. Ideally we still want to look at these peptides to make sure we are not missing important ones. 

This script pulls the Manual Review comments of these peptides and analyze them whether these were worthwhile peptides. 

**NOTE: only joining the comments after the Manual Review since the ITB Evaluation may be changed after manual revision**

# Combine predicted and true evaluation from these files
```{r}
fnames <- c("5120-16", "5120-17", "5120-18", "5120-19") # patient ID
fnames_new <- c("5120-16", "5120-17", "5120-18", "5120-19", "5120-39", "CTEP-10146-MD017-0052", "G110_Region1", "G110_Region2", "G113_Region1")
pred_dir <- here("output", "08-1_threshold_adj.Rmd") # random forest predictions
manual_review_dir <- here("data", "files_for_prediction", "with_manual_review_comments") # the itb file after the manual review process
```

Function to merge itb reviewed file, true Evaluation, and predicted Evaluations, probabilities, manual review comments, and final Evaluation decision together into one file for analysis

```{r}
merge_pred_eval_comments <- function(patient_id, pred_dir, manual_review_dir, outdir) {
  # Locate and read prediction file (that has prediction, probabilities, and actual Evaluations)
  predict_files <- list.files(pred_dir, full.names = TRUE)
  predict_file_path <- grep(paste0(patient_id, ".*predict&trueEval_newThreshold.*\\.tsv$"), predict_files, value = TRUE)
  
  if (length(predict_file_path) == 0) {
    stop(paste("No prediction file found for patient:", patient_id))
  }
  # Check if exactly one prediction file is found
  if (length(predict_file_path) != 1) {
    stop(paste("Error: Found", length(predict_file_path), "prediction files for patient ID:", patient_id,
               "\nPlease ensure there is exactly one matching prediction file."))
  }
  
  predict_file <- read.delim(predict_file_path)
  
  # Locate and read evaluation file
  manual_review_files <- list.files(manual_review_dir, full.names = TRUE)
  manual_review_file_path <- grep(patient_id, manual_review_files, value = TRUE)
  
  if (length(manual_review_file_path) == 0) {
    stop(paste("No evaluation file found for patient:", patient_id))
  }
  # Check if exactly one manual review file is found
  if (length(manual_review_file_path) != 1) {
    stop(paste("Error: Found", length(manual_review_file_path), "manual review files for patient ID:", patient_id,
               "\nPlease ensure there is exactly one matching manual review file."))
  }
  
  manual_review_file <- read.xlsx(manual_review_file_path) %>% 
    rename(final_Evaluation = Evaluation)
  # columns of comments: pVAC.Review.Comments, IGV.Review.Comments
  
  # Merge prediction and evaluation data
  merged_df_export <- left_join(
    predict_file,
    manual_review_file %>% select(
      "ID",
      "final_Evaluation",
      "pVAC.Review.Comments",
      "IGV.Review.Comments"
    ),
    by = "ID"
  ) %>%
    mutate(patient_id = patient_id) %>%
    mutate(Pos = if (is.integer(Pos))
      as.character(Pos)
      else
        Pos) %>% # make sure Pos is always type chr, sometimes it gets converted to int and messes up merging in the do.call step 
    mutate(Ref.Match = if (!is.character(Ref.Match))
      as.character(Ref.Match)
      else
        Ref.Match)
  
  # Export merged data frame as excel format
  out_name = paste0(patient_id, "_", "predict&trueEval&manualComments.xlsx")
  write.xlsx(merged_df_export, 
             file = file.path(outdir, out_name)
  )
  
  # Return the merged data frame
  return(merged_df_export)
}

# Process all files
all_files <- lapply(fnames_new, merge_pred_eval_comments, pred_dir, manual_review_dir, outdir)

# Combine all data frames
all_df <- do.call(bind_rows, all_files)

# Export the combined data frame
write.xlsx(all_df, file.path(outdir, "all_predict&trueEval&manualComments.xlsx"), rowNames = FALSE)
```

Filter for peptides which are predicted to be "Reject" but are actually marked as "Review"

```{r}
problematic_rejects <- all_df %>% 
  filter(Evaluation_pred == "Reject" & Evaluation == "Review")

# Export the combined data frame
write.xlsx(problematic_rejects, file.path(outdir, "problematic_rejects.xlsx"), rowNames = FALSE)
```

