---
title: "Summary statistics"
author: "Jennie Yao"
date: "2025-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(here)
library(data.table)
```

Output directory
```{r set output directory}
outdir <- here("output", "summary.Rmd")
fs::dir_create(outdir)
```


Goal: 
* To summarize statistics for training and testing sets


OUTCOME: 
Output file: 


Read metadata and imputed data (the step right before training and testing)
```{r read metadata}
meta <- fread(here("output", "02_make_meta2.Rmd", "metadata_count_purity.csv"))
df_imputed <- readRDS(here("output", "05_cleaning_imputation.Rmd", "df_imputed.rds"))
```

# For knowing how many peptides are in each trial (training + test set)
This part was added for making the CGC 2024 poster table. I wanted to know how many peptides were in each trial. 

A list of patient_id with the trial name
```{r extract patient id with the trial name they are from}
patient_id_trial <- meta %>% 
  filter(include_ML == "Yes") %>% 
  select(patient_id, trial, tumor_type)
```

```{r}
# add trial names to df_imputed (extra column)
# Excluding "Review" peptides since they were excluded from model training and testing
# "Pending" are grouped into "Rejects"
df_trial <- df_imputed %>% 
  filter(Evaluation != "Review") %>% 
  mutate(Evaluation = ifelse(Evaluation == "Accept", 1, 0)) %>% 
  full_join(., patient_id_trial, by = "patient_id")
#df_trial <- full_join(df_imputed, patient_id_trial, by = "patient_id")


# Group by the 'trial' column and calculate the number of unique patient_id, the number of rows in each group, the number of Accepts, the number of Rejects
result <- df_trial %>%
  group_by(trial, tumor_type) %>%
  summarise(
    total_patients = n_distinct(patient_id),
    total_peptides = n(), 
    count_accept = sum(Evaluation == 1, na.rm = TRUE),
    count_reject = sum(Evaluation == 0, na.rm = TRUE)
  )

# Print the result
print(result)
```